Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-28 article.tex
Fix formula for trajectory distance

In order to perform clustering, we define a distance between trajectories.
A frequency trajectory of a mutation $X$ is characterized by a series of frequency values $f(t)$ at time point  $t\in T$.
We define $S(t)$ as the strains that carry mutation $X$ at date $t$.
With this notation, $f(t)$ is the ratio of the number of elements in $S(t)$ to the total number of strains at date $t$.
Let us now consider two frequency trajectories $X_1$ and $X_2$. \\
We define the distance $d(X_1,X_2)$ between these two trajectories based on the average similarity of the strains $S_1$ and $S_2$ that compose them:
$$ d(X_1,X_2) = \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$
where $T_1\cap T_2$ is the time interval where both trajectories are active, and $\vert\cdot\vert$ denotes the number of elements of a set.
The quantity summed corresponds to the Jaccard index between strains composing $X_1$ and $X_2$ at a given date.
It is 1 if the two trajectories share exactly the same strains for this date, and 0 if they share no strain at all.
This leads to the two following properties of $d$:
\begin{itemize}
	\item if $d(X_1,X_2)=0$, then $X_1$ and $X_2$ represent the same frequency trajectory. The mutations $x_1$ and $x_2$ that they correpond to always appear on the same strains and are totally linked.
	\item if $d(X_1,X_2)=1$, then $X_1$ and $X_2$ can be considered completely independent. This can be the case if the two trajectories do not occur at the same dates, *i.e.* $\vert T_1\cap T_2\vert=0$, or if their respective mutations are never present on the same genomes.
\end{itemize}

>>>>

In order to perform clustering, we define a distance between trajectories.
A frequency trajectory of a mutation $X$ is characterized by a series of frequency values $f(t)$ at time point  $t\in T$.
We define $S(t)$ as the strains that carry mutation $X$ at date $t$.
With this notation, $f(t)$ is the ratio of the number of elements in $S(t)$ to the total number of strains at date $t$.
Let us now consider two frequency trajectories $X_1$ and $X_2$. \\
We define the distance $d(X_1,X_2)$ between these two trajectories based on the average similarity of the strains $S_1$ and $S_2$ that compose them:
$$ d(X_1,X_2) = 1 - \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$
where $T_1\cap T_2$ is the time interval where both trajectories are active, and $\vert\cdot\vert$ denotes the number of elements of a set.
The quantity summed corresponds to the Jaccard index between strains composing $X_1$ and $X_2$ at a given date.
It is 1 if the two trajectories share exactly the same strains for this date, and 0 if they share no strain at all.
This leads to the two following properties of $d$:
\begin{itemize}
	\item if $d(X_1,X_2)=0$, then $X_1$ and $X_2$ represent the same frequency trajectory. The mutations $x_1$ and $x_2$ that they correpond to always appear on the same strains and are totally linked.
	\item if $d(X_1,X_2)=1$, then $X_1$ and $X_2$ can be considered completely independent. This can be the case if the two trajectories do not occur at the same dates, *i.e.* $\vert T_1\cap T_2\vert=0$, or if their respective mutations are never present on the same genomes.
\end{itemize}
